The C. elegans gustatory receptor homolog LITE-1 is a chemoreceptor required for diacetyl avoidance

Version 1

Abstract

Abstract

The nematode C. elegans does not have eyes but can respond to aversive UV and blue light stimulation and even distinguish colours. The gustatory receptor homolog LITE-1 was identified in forward genetic screens for worms that failed to respond to blue light stimulation. When LITE-1 is expressed in body-wall muscles, it causes contraction in response to blue light suggesting that LITE-1 is both necessary and sufficient for blue light response. Here we show that in addition to light avoidance, LITE-1 is also required for worms’ avoidance of high concentrations of diacetyl, an odorant that is attractive at low concentrations. Like blue light, diacetyl causes muscle contraction in transgenic worms engineered to express LITE-1 in body-wall muscles. These data are consistent with a direct chemoreceptor function for LITE-1 which would make it a multimodal sensor of aversive stimuli.

Introduction

Sensory perception is fundamental to how organisms interact with their surroundings. Animals, including Caenorhabditis elegans, detect and respond to a wide range of environmental cues such as light and odorants. Despite having a compact nervous system of only 302 neurons, C. elegans exhibits sensory capabilities that allow it to effectively navigate and adapt to its environment (Bargmann, 2006; Cook et al., 2019). Odorants such as volatile organic compounds are detected via specialised olfactory receptor neurons. These neurons express specific receptors that bind to specific molecules and trigger distinct behavioural responses (Bargmann, 2006; Bargmann et al., 1993; Bargmann & Horvitz, 1991). Diacetyl, a volatile organic α-diketone found in C. elegans’ natural environment elicits opposing behaviours depending on its concentration. At low concentrations, diacetyl is sensed by the olfactory receptor ODR-10, promoting attraction (Itskovits et al., 2018; Sengupta et al., 1996; Zhang et al., 1997). In contrast, higher concentrations induce an avoidance response, thought to be mediated by the receptor SRI-14 (Taniguchi et al., 2014).

In addition to chemical cues, C. elegans also detects and responds to light, despite lacking eyes or classical photoreceptors. This light sensitivity is mediated by LITE-1, a member of the Gustatory Receptor family (Montell, 2009). LITE-1 was first identified through genetic screens for mutants defective in light avoidance behaviour (Edwards et al., 2008). Unlike canonical photoreceptors such as rhodopsins or opsins, LITE-1 shares little sequence similarity with known light sensing proteins (Edwards et al., 2008; Gong et al., 2016; Liu et al., 2010). Further underscoring its uncanonical nature, LITE-1 adopts an inverted membrane topology that is unlike classical G protein-coupled receptors (GPCRs) (Gong et al., 2016).

LITE-1 phototaxis behaviour is driven by multiple light responsive sensory neurons, including ASJ, ASH and ASK (Liu et al., 2010; Ward et al., 2008; Zhang et al., 2022). Signal transduction occurs via G protein signalling, upregulation of cGMP levels and activation of cyclic nucleotide-gated (CNG) channels, leading to a rapid avoidance response (Edwards et al., 2008; Liu et al., 2010; Ward et al., 2008). Recent studies suggest that LITE-1 may function as a light activated ion channel with a putative ligand binding pocket that is activated by ultraviolet and blue light via key tryptophan residues (Edwards et al., 2008; Gong et al., 2016; Hanson, Scholüke, et al., 2023; Ward et al., 2008).

Here we show that beyond its role in photoreception, LITE-1 is also required for the avoidance of high concentrations of diacetyl and that diacetyl induces muscle contraction in transgenic worms expressing LITE-1 in body-wall muscles. Our findings support the notion that LITE-1 functions as a multifunctional receptor involved in both light and chemical sensing.

Results

LITE-1 is required for diacetyl and blue light avoidance

A recent screen of a panel of worm strains that collectively contained knockouts of all non-essential GPCRs identified a triple mutant that failed to avoid high concentrations of diacetyl (Pu et al., 2023). One of the mutated genes was lite-1. We performed a chemotaxis assay to test single mutants of several loss-of-function alleles of lite-1 and found they all failed to avoid high concentrations of diacetyl (Fig. 1A and 1B) while remaining attracted to lower concentrations (Fig. 1C). As expected, these lite-1 mutants no longer respond to blue light stimulation (Fig. 1D) (Edwards et al., 2008).

(A) Schematic of chemotaxis assay setup. (B) Chemotaxis index of wild-type animals and mutants with different alleles of lite-1 that have defects in diacetyl response. P-values are from unpaired two-tailed t-tests, n = 4 biological replicates, each an average of 4 plates. (C) Chemotaxis index of wild-type and lite-1 (yum2880) animals with different diacetyl concentrations (5 μl). P-values are from unpaired two-tailed t-tests, n = 3 biological replicates, each an average of 4 plates. (D) Time series of wild-type (blue line) animals and mutants with different alleles of lite-1 (orange line) when stimulated with blue light (blue shaded box). n = 3 biological replicates, each an average of 12 wells.

Chemosensory neurons ADL, ASK and ASH are involved in avoidance of diacetyl

To identify neurons involved in the response to high concentrations of diacetyl, we used calcium imaging to record from 11 pairs of sensory neurons in worms exposed to a pulse of diacetyl (Fig. 2A). In lite-1 mutants, the LITE-1-expressing sensory neurons ADL and ASK showed a clear defect in diacetyl response, whereas another LITE-1-expressing neuron ASH showed a normal response (compared to WT animals, Fig. 2B). The response of the other tested neurons was not significantly different between mutant and WT animals (Fig. 2B). The normal response to diacetyl in ASH in the absence of LITE-1 is likely due to its expression of SRI-14 which has previously been shown to respond weakly to high concentration of diacetyl (Taniguchi et al., 2014). We also recorded from unc-13 mutants which are impaired in synaptic activity and found little effect on the ADL and ASK responses (Fig. 2C) (Richmond et al., 1999). This is consistent with direct sensation in these neurons although neural signalling through gap junctions or via neuromodulation by other receptors could not be ruled out.

(A) Confocal image of labelled sensory neurons and calcium imaging traces showing the responses of three neurons when exposed to high concentrations of diacetyl. (B) Neuronal activity of wild-type controls with lite-1 mutants. (C) Neuronal activity of wild-type controls to unc-13 mutants. Grey shaded box denotes the presence of diacetyl. Shaded area around the activity dynamics denotes standard error.

LITE-1 is sufficient for diacetyl response in other cell types

To test whether LITE-1 is sufficient for a response to high diacetyl concentrations, we used a transgenic strain that expresses LITE-1 in body-wall muscles (Edwards et al., 2008). Worms ectopically expressing LITE-1 in body muscle cells evoke muscle contraction and shortening of body length in response to blue light stimulation (Edwards et al., 2008). When these worms are exposed to a 1:50 dilution of diacetyl, they showed a rapid and near complete paralysis whereas wild-type animals continue to swim (Fig. 3A and 3B). These worms also showed significant contraction in average body length, contracting by ∼11% after diacetyl exposure (Fig. 3C). These results suggest that LITE-1 is directly activated by diacetyl although it remains possible that it is a diacetyl metabolite that activates LITE-1 given that the full contraction develops over a minute time scale. Consistent with a ligand receptor relationship, molecular docking predicts low micromolar affinity of diacetyl binding to LITE-1 in the same binding pocket that was recently identified as a putative chromophore binding site (Fig. 3D) (Hanson, Scholüke, et al., 2023).

(A) Percent of worms that are swimming at two minutes after exposure to buffer untreated or to diacetyl. Wild-type animals continue swimming in buffer and diacetyl (WT, left) whereas worms expressing LITE-1 in body wall muscles (Pmyo-3::LITE-1, right) are mostly paralysed when treated with diacetyl. (B) Sample midline trajectories from wild-type and Pmyo-3::LITE-1 animals showing difference in motion (each trajectory shows 6 seconds of swimming). (C) Worm body length over time after treatment. Wild-type animals do not show a significant length difference in either buffer or diacetyl treatments (left) whereas worms expressing LITE-1 in body wall muscles show a significant contraction in response to diacetyl (right; insert showing the average contraction). P-values from paired two-tailed t-test, n ≥ 12 worms. (D) Molecular docking of diacetyl into the AlphaFold-predicted structure of LITE-1 with predicted micromolar affinity.

2,3-pentanedione also interacts with LITE-1

We next tested whether lite-1 is required for avoidance of other high concentration odorants. Of the seven odorants tested, lite-1 mutants showed a defective response to 2,3-pentanedione, a chemical structurally related to diacetyl (Fig. 4A). The effect was weaker than that observed for diacetyl, with worms still avoiding high concentrations of 2,3-pentanedione but more weakly than wild-type animals. Supporting this observation, molecular docking predicts interaction between 2,3-pentanedione and LITE-1 putative binding pocket (Fig. 4B). Like diacetyl, 2,3-pentanedione also evokes paralysis and muscle contraction of worms expressing LITE-1 in muscle cells (Fig. 4C and 4D).

(A) Chemotaxis index of wild-type animals and lite-1 mutants for known odorants. P-values are from unpaired two-tailed t-tests, n = 4 biological replicates each an average of 4 plates. (B) Molecular docking of 2,3-pentanedione into the AlphaFold-predicted structure of LITE-1 with predicted micromolar affinity. (C) Percent of worms that are swimming at two minutes after exposure to buffer untreated or to 2,3-pentanedione. Wild-type animals continue swimming in buffer and 2,3-pentanedione (WT, left) whereas worms expressing LITE-1 in body wall muscles (Pmyo-3::LITE-1, right) are mostly paralysed when treated with 2,3-pentanedione. (D) Worm body length over time after treatment. Wild-type animals do not show a significant length difference in either buffer or 2,3-pentanedione treatments (left) whereas worms expressing LITE-1 in body wall muscles show a significant contraction in response to 2,3-pentanedione (right; insert showing the average contraction). P-values from paired two-tailed t-test, n ≥ 12 worms.

Discussion

LITE-1 is best known for its role in light sensation in C. elegans, but how it senses light is not yet clear. In particular, an indirect mechanism in which light produces photoproducts that activate the channel has not been ruled out and a priori has always seemed plausible given that LITE-1 is similar to gustatory receptors that are known to sense chemicals (Montell, 2009). A direct diacetyl response would make a photoproduct-based mechanism more plausible. The identification of a second structurally related compound, 2,3-pentanedione may help in narrowing down photoproducts with similar moieties. Direct and photoproduct light sensing are not mutually exclusive. Based on LITE-1’s predicted structure and its absorption spectrum, it has recently been proposed that UV may be directly sensed while the blue light response may be mediated through a photoproduct or an unidentified chromophore (Hanson et al., 2023).

While our study primarily focused on chemosensory neurons, it remains possible that non-chemosensory neurons expressing LITE-1 may contribute to diacetyl detection or modulating the behavioural response. Indeed, a recent study demonstrated that the LITE-1-expressing interneuron AVG response to blue light stimulation by releasing the neuropeptide NLP-10 to mediate the escape behaviour (Dunkel et al., 2025). Nevertheless, phototransduction and chemotransduction appear to involve shared sensory neurons, such as ASK and ASH, suggesting a potential overlap in their underlaying signalling mechanisms (Zhang et al., 2022). It is possible that LITE-1 chemosensory response may rely on components necessary for LITE-1 photosensory function, such as G proteins (GOA-1, GPA-3), guanylate cyclase (DAF-11, ODR-1), CNG channels (TAX-2,4) and phosphodiesterase (PDE-1,2,5) (Liu et al., 2010; Ward et al., 2008). Furthermore, the use of lite-1 point mutants that affect specific LITE-1 function such as light sensing, channel gating or binding pocket could further elucidate LITE-1 mechanisms (Gong et al., 2016; Hanson, Scholüke, et al., 2023).

We have shown that LITE-1 is required for the avoidance of high concentrations of diacetyl and a related compound 2,3-pentanedione. Furthermore, we have shown that LITE-1 is sufficient to make muscle cells responsive to diacetyl and 2,3-pentanedione, consistent with a direct response. Diacetyl is an ecologically relevant molecule present in C. elegans’ natural environment that has been hypothesised to be involved in prey-finding since low concentrations produced by lactic acid bacteria on rotting fruit is attractive (Choi et al., 2016). Avoidance of high concentrations of diacetyl may therefore also be ecologically relevant. In this case, LITE-1 has likely evolved to be a multimodal receptor of aversive stimuli akin to other multimodal receptors such as TRPV1, which responds to heat, capsaicin, and acidic pH (Zhang et al., 2023).

Material and methods

Bacterial and worm strains maintenance

The E. coil wild-type isolate OP50 was sourced from Caenorhabditis Genetics Center (CGC) and maintained on lysogeny broth (LB) agar under standard laboratory conditions. The following worm strains were used in this study, N2; wild-type isolate (CGC), CHS1610; lite-1(yum2880) (Pu et al., 2023), KG1180; lite-1(ce314) (Edwards et al., 2008), KG1272; cels37[myo-3prom::lite-1::GFP] (Edwards et al., 2008), TQ8245; lite-1(xu492) (Zhang et al., 2020), ZAS280; In[osm-6::GCaMP3, osm-6::ceNLS-mCherry-2xSV40NLS] (Iwanir et al., 2019), ZAS371; unc-13(s69) In[osm-6::GCaMP3, osm-6::ceNLS-mCherry-2xSV40NLS] (Bokman, Kalij, et al., 2024; Bokman, Pritz, et al., 2024), ZAS526; lite-1(ce314) In[osm-6::GCaMP3, osm-6::ceNLS-mCherry-2xSV40NLS], were cultured on Nematode Growth Medium seeded with E. coli (OP50) as previously described (Brenner, 1974). Worm synchronization was performed by bleaching unsynchronised gravid adults (Barlow, 2019).

Chemotaxis assay

The chemotaxis assays were conducted at 20°C in 9 cm plates filled with 10 ml of agar (1.7% agar, 1 mM CaCl2, 1 mM MgSO4 and 25 mM K2HPO4 pH 6). Two parallel lines, 1 cm apart, were drawn on the middle of the plates which define the area to place the animals. Two odorant spots were designated on one side of the plate, while two control spots were positioned on the opposite side (Fig. 1A). 1 μl of 1 M NaN3 was added to all four spots to immobilise the animals. Different volumes of each odorant: diacetyl (5 μl), 2,4,5-trimethylthiazole (7.5 μl), benzaldehyde (2.5 μl), isoamyl alcohol (7.5 μl), 2,3-pentanedione (10 μl), 2-nonanone (1 μl), 1-octanol (2.5 μl), 2-butanone (10 μl) and equal volume of ethanol (control) were spotted onto the respective dots. Synchronized day-one adults were washed 3 times in chemotaxis buffer (1 mM CaCl2, 1 mM MgSO4 and 25 mM K2HPO4 pH 6) before being transferred onto the centre of the assay plate where the assay will last for 1 hr at 20 °C. The chemotaxis plates were then stored at 4 °C before counting. The chemotaxis index was calculated as (number of animals on the odorant side – number of animals on the ethanol side) / (sum of all animals on both sides).

Paralysis assay

The paralysis assays were conducted at 20°C in 6-well plates with 4 ml of agar (1.7% agar, 1 mM CaCl2, 1 mM MgSO4 and 25 mM K2HPO4 pH 6). Synchronized day-one adults were washed 3 times in chemotaxis buffer (1 mM CaCl2, 1 mM MgSO4 and 25 mM K2HPO4 pH 6) before being transferred onto the centre of the 6-well plate. 1 ml of chemotaxis buffer containing diacetyl (1:50) was then added into each well and videos were immediately acquired at 25 frames per second using a shutter speed of 25 ms and a resolution of 12.4 μm/pixel.

Blue light stimulation assay

The blue light stimulation assays were conducted at 20°C in 24-well plates containing agar (1.7% agar, 1 mM CaCl2, 1 mM MgSO4 and 25 mM K2HPO4 pH 6). Synchronized day-one adults were transferred onto the centre of the 24-well plate pre-seeded with OP50. The plates were then transferred onto the multi-camera tracker for 30 min to allow acclimatization to the tracker room before imaging. Videos were acquired at 25 frames per second using a shutter speed of 25 ms and a resolution of 12.4 μm/pixel. Videos were taken in the following order: 300 s pre-stimulus, blue light stimulus with one 10 s pulse of blue light at 60 s and a 100 s post-stimulus recording. Videos were then processed with Tierpsy Tracker as previously described (Barlow et al., 2022; Javer et al., 2018).

Calcium imaging

Young adult worms were inserted into a microfluidic chip and anesthetized with 10 mM levamisole and allowed to habituate for 10 mins as previously described (Chronis et al., 2007). Recordings began with 1 min of light habituation, after which the stimulus was presented for 1 min The stimulus used was 1:5 diacetyl in CTX buffer (5 mM KH2PO4/K2HPO4 pH 6, 1 mM CaCl2 and 1 mM MgSO4). Recordings were made using a Nikon A1R + confocal laser scanning microscope with a water immersion × 40 (1.15NA) objective controlled by the Nikon NIS-elements software. Z-axis resolution was 0.5-0.8 μm. Acquisition rate was ∼1.5 volumes/second, interpolated to a final framerate of 2 Hz.

Neuron identification and signal extraction

Neuron segmentation and tracking was done on the nuclear mCherry signal using an algorithm developed by (Toyoshima et al., 2016). Neurons were identified visually based on anatomical positions. Neurons that could not be unambiguously identified were removed from analysis. Calcium readings were taken from within 0.9 of the radius of the segmented nuclear sphere. Signal intensity was normalized by baseline activity, defined as the lowest point of a 10-frame running average. All data analyses were performed using in-house MATLAB scripts (Pritz et al., 2023).

Molecular docking

Molecular docking was performed on the LITE-1 tetramer structure using DynamicBind (Hanson, Scholuke, et al., 2023; Lu et al., 2024). Additional rounds of docking on top ranked poses were performed with Gnina (McNutt et al., 2021).

eLife Assessment

Avoidance of UV and blue light by the nematode C. elegans is mediated by the unusual transmembrane protein LITE-1, a non-canonical photoreceptor. In this valuable work, the authors provide convincing evidence that LITE-1 function is also required for avoidance of very high concentrations of the food-associated cue diacetyl, suggesting that it may also function as a diacetyl chemoreceptor. While the evidence for this idea is incomplete, these intriguing findings suggest an unexpected complexity in the function of this unusual photoreceptor.

Reviewer #1 (Public review):

Summary:

This paper describes an interesting phenotype of C. elegans lite-1 mutants. Previous work showed that lite-1 mutants lose a violet/blue light avoidance response. The authors show here that lite-1 mutants also show a defect in negative diacetyl chemotaxis. While wild-type worms avoid diacetyl at high concentrations, lite-1 mutants are instead *attracted* to it. The authors go on to perform Ca2+ imaging in sensory neurons and find that ADL and ASK neurons show altered Ca2+ responses to diacetyl in lite-1 mutants, suggesting LITE-1 is required for these responses. As unc-13 mutants with defective synaptic transmission show similar diacetyl Ca2+ responses as wild-type, this suggests these neurons respond cell autonomously to diacetyl. However, whether lite-1 also acts cell-autonomously is not discussed. Indeed, because unc-13 and lite-1 mutants show different ADL and ASK Ca2+ responses, it seems the diacetyl response regulated by LITE-1 is likely acting outside of those cells. An interesting result that is not commented on is the switching of the valence of the ASK Ca2+ response in lite-1 mutants. ASK neurons still respond to diacetyl, but instead of a strong increase in Ca2+, diacetyl appears to drive it strongly lower. This may be consistent with the switch in valence in the diacetyl chemotaxis assay. It also argues against the idea that LITE-1 is a low-affinity diacetyl receptor that drives avoidance or the Ca2+ responses in ASK, since it is still present in lite-1 mutants. The authors then use a strain that expresses LITE-1 in the body wall muscles and show this expression is sufficient to engender them with sensitivity to diacetyl, as measured through altered swimming and hypercontractility. The authors interpret this result as LITE-1 may act as a diacetyl receptor. The authors test whether a structurally similar molecule, 2,3-pentanedione, shows similar effects, and they find it does. Alpha-fold modeling and molecular docking analysis show where diacetyl might bind to the LITE-1 protein. They then test whether lite-1 mutants show chemotaxis defects to other molecules, as seen with diacetyl. Generally, they find that the observed diacetyl responses are unique, although lite-1 mutants do lose their avoidance response to 2,3-pentanedione. However, unlike the acquisition of diacetyl attraction in lite-1 mutants, 2,3 pentanedione avoidance is *lost*; it is not switched to attraction. Overall, I felt the description of the results and their implications could have been more in-depth. Further, the evidence that LITE-1 is a chemoreceptor itself, rather than acting in some way to shape chemoreceptor responses (via light or otherwise), remains unclear, as conceded by the authors.

Strengths:

Overall, the study follows up on an interesting and useful result. The experiments as presented are generally well-conceived and performed. The authors use a variety of behavioral and imaging approaches to test how LITE-1 mediates diacetyl avoidance.

Weaknesses:

The study is missing experiments needed to resolve whether LITE-1 is doing what they propose. The evidence that LITE-1 is a diacetyl receptor is lacking support since lite-1 mutants have their avoidance and calcium responses flipped, which would not be expected if it were acting solely as an avoidance receptor. Presumably, the authors are concluding that the attractive response that is left in the lite-1 mutant is mediated by ODR-10, but that experiment is not shown. Similarly, the authors concede that "the use of lite-1 point mutants that affect specific LITE-1 function, such as light sensing, channel gating, or binding pocket, could further elucidate LITE-1 mechanisms." This reviewer agrees, and such experiments designed to localize diacetyl binding site(s) would be necessary to conclude definitively that LITE-1 is a diacetyl receptor. The body wall muscle assay used or some other heterologous experimental system could work for such a structure-function analysis. A concern is whether the extensive number of LITE-1 point mutants described in the literature affect cell surface expression vs. receptor function, which might complicate the interpretation of a result showing loss of diacetyl responses.

Reviewer #2 (Public review):

Summary:

Koh and colleagues investigate the broader sensory role of LITE-1, a gustatory receptor previously linked to UV light detection in C. elegans. Their study explores whether LITE-1 also mediates avoidance of specific chemical stimuli-namely, high concentrations of diacetyl and 2,3-pentanedione. They show that LITE-1 is required in the ADL and ASK neurons for calcium responses to diacetyl, and that its expression in body-wall muscles is sufficient to trigger hypercontraction upon odorant exposure. Molecular docking suggests both odorants may directly bind to LITE-1 with micromolar affinity. These findings suggest LITE-1 may act as a multimodal receptor for both light and chemical stimuli.

Strengths:

(1) Methodological Precision: The study is technically strong, with well-executed calcium imaging and quantitative behavioral assays that clearly show neural and muscular responses to chemical stimuli.

(2) Novelty and Scope: The work presents a compelling case for LITE-1 functioning as a multimodal sensor, which is an intriguing expansion of its known role.

(3) Potential Impact: If validated, the findings could significantly advance the understanding of sensory integration in C. elegans, and the tools developed may be broadly useful to the research community.

(4) Relevance to the Field: The study adds to evidence that C. elegans uses non-canonical sensory pathways and may inspire further exploration of multimodal receptor functions in other systems.

Weaknesses:

(1) Lack of Rescue Experiments: The absence of rescue experiments makes it difficult to definitively link the observed phenotypes to loss of lite-1.

(2) Single Loss-of-Function Approach: The reliance on a single genetic mutant limits interpretability. Additional strategies such as RNAi (e.g., neuron-specific knockdown) would provide stronger evidence.

(3) Unclear Neuronal Contribution: While calcium responses in ADL and ASK are reduced, it's unclear which neuron(s) are necessary for behavioral avoidance. Cell-specific rescue or knockdown experiments are needed.

(4) Unvalidated Docking Data: The molecular docking predictions lack experimental validation. Site-directed mutagenesis would be needed to support claims of direct interaction.

(5) Limited Odorant Specificity Testing: Docking analysis does not include non-binding odorants, making it difficult to assess binding specificity.

(6) Incomplete Quantification: Some calcium imaging results (e.g., in AWA neurons of unc-13 mutants) lack statistical comparisons, which limits their interpretive value.

Reviewer #3 (Public review):

In this work, Brown and colleagues report that the photosensor protein LITE-1 of the nematode C. elegans may also be a chemosensor that can be activated by high concentrations of the compound diacetyl. LITE-1 was described as a putative ion channel of the gustatory receptor family, which is mainly constituted by insect odorant receptors. These form tetrameric ion channels that can be activated by odorants. Specificity is achieved by forming heteromeric channels from three copies of the odorant receptor co-receptor (ORCO) and another subunit that resembles ORCO in the pore-forming C-terminus, but brings in a binding site for the respective odorant. LITE-1 has a very similar structure, according to Alphafold3 predictions, and also carries a binding pocket. In LITE-1, this was proposed to be occupied by a light-absorbing molecule that activates the channel when a photon is absorbed. Alternatively, compounds generated by absorption of high-energy photons may be formed in vivo and bound by the LITE-1 binding pocket. Koh et al. now demonstrate that another, non-light-activated compound, diacetyl, at high concentrations, can activate cells expressing LITE-1. Such (chemosensory) cells are also responsible for the avoidance of high concentrations of diacetyl. LITE-1 activation in excitable cells, i.e, muscles, causes strong body contraction and paralysis, and the authors show that this is also the case when diacetyl is presented. The authors further present molecular docking studies showing that diacetyl could occupy the binding pocket of LITE-1. Last, they show that another compound chemically resembling diacetyl, i.e., 2,3-pentanedione, can also induce avoidance in a LITE-1 dependent manner, though not as potently.

The data are intriguing, and the demonstration of LITE-1 being a diacetyl chemosensor is interesting. Yet, there are a few questions arising that the authors should address.

The authors identified mutants lacking diacetyl responses. In their chemotaxis assay (Figures 1A, B), they show that lite-1 mutants do not avoid high concentrations of diacetyl. However, the animals actually showed attraction, as the chemotaxis index was positive. If the lite-1 animals were insensitive, they should be indifferent, and the chemotaxis index should be close to zero. This means, other neurons contribute to the diacetyl response, and the result of these neurons being activated means/remains attraction? If so, the authors need to rule out any effects of these neurons on the effects they attribute to LITE-1 in the other assays.

The effect of diacetyl on muscle cells (Figure 3C) is pretty rapid, i.e., already during 1 minute after application, the animals are almost maximally contracted. How fast is it really? Can the authors provide a time course with more time points during the first minute? This is a relevant question, as the compound would have to either pass the worm cuticle or enter through the gut and diffuse through the body to reach the muscle cells. Can one expect this to occur within (less than) a minute?

In this context, the authors need to rule out that other mechanisms may be at play. E.g., diacetyl may be immediately sensed by ciliated chemosensory neurons that might release a signaling molecule that leads to activation of LITE-1 in muscles, or that sensitizes it somehow, responding to light used for filming animals. The authors should repeat this assay in a lite-1 mutant background. Furthermore, the authors tested unc-13 mutants to rule out indirect effects on the neurons recorded. Likewise, they should eliminate neuropeptide signaling via unc-31 mutants (a recent paper cited by the authors showed involvement of neuropeptide signaling in LITE-1-mediated light avoidance behavior). Last, to demonstrate that effects are not indirect in response to chemosensory neurons, the authors should repeat the contraction or swimming assay in a tax-4 mutant, which largely lacks chemosensation. This also applies to the chemotaxis assay. Animals should exhibit a chemotaxis index to diacetyl of zero, then.

Does diacetyl activate other neurons expressing LITE-1? A number of cells express LITE-1 at high levels, which the authors have not tested (they restricted their analyses to chemosensory neurons). This is important to address because it leaves the possibility that LITE-1 requires a specific partner only present in these chemosensory neurons to detect diacetyl. This partner would have to be present also in muscles, where diacetyl could activate ectopically expressed LITE-1. According to CeNGEN scRNAseq data, cells expressing LITE-1 can be identified. The ADL and ASH neurons actually come up only at the lowest threshold, so some of the other cells showing much higher levels of LITE-1 mRNAs, i.e., AVG, ALM, PLM, ASG, PHA, PHB, AVM, RIF, or some pharyngeal neurons, should be tested. ASG was among the cells the authors recorded from, but this neuron did not show a response.

The authors need to show that diacetyl responses of ADL and/or ASK can be rescued by expressing LITE-1 specifically in these neurons in a lite-1 mutant background.

Molecular docking studies are not described in detail. How was this done? Diacetyl is a very small molecule. How well can docking algorithms assess this at all? Did the authors preselect the binding pocket, or did the algorithm sample the entire molecular surface of the LITE-1 model and end up with the binding pocket? The latter would be very convincing. The authors should provide control docking experiments with other molecules that caused avoidance in their hands (i.e. benzaldehyde, 2,4,5,trimethlythiazole, isoamyl alcohol, nonanone, octanone), but did not activate LITE-1. Also, they should try docking molecules related to diacetyl, and if there are some that do not dock under the same conditions, such molecules should be used in a behavioral experiment. Ideally, they should also not activate LITE-1. Examples could be, e.g., diacetyl monoxime or 2,4-pentanedione.

Last, the authors should provide a PDB file with the docked diacetyl to allow readers to assess the binding for themselves. Since a large number of mutations of LITE-1 have been reported, it may be that amino acids shown to be essential for LITE-1 function are also required for diacetyl binding. If so, this could be backed up with an experiment.

Author response:

The following is the authors’ response to the original reviews.

Public Reviews:

Reviewer #1 (Public review):

Summary:

This paper describes an interesting phenotype of C. elegans lite-1 mutants. Previous work showed that lite-1 mutants lose a violet/blue light avoidance response. The authors show here that lite-1 mutants also show a defect in negative diacetyl chemotaxis. While wild-type worms avoid diacetyl at high concentrations, lite-1 mutants are instead *attracted* to it. The authors go on to perform Ca2+ imaging in sensory neurons and find that ADL and ASK neurons show altered Ca2+ responses to diacetyl in lite-1 mutants, suggesting LITE-1 is required for these responses. As unc-13 mutants with defective synaptic transmission show similar diacetyl Ca2+ responses as wild-type, this suggests these neurons respond cell autonomously to diacetyl. However, whether lite-1 also acts cell-autonomously is not discussed. Indeed, because unc-13 and lite-1 mutants show different ADL and ASK Ca2+ responses, it seems the diacetyl response regulated by LITE-1 is likely acting outside of those cells. An interesting result that is not commented on is the switching of the valence of the ASK Ca2+ response in lite-1 mutants. ASK neurons still respond to diacetyl, but instead of a strong increase in Ca2+, diacetyl appears to drive it strongly lower. This may be consistent with the switch in valence in the diacetyl chemotaxis assay. It also argues against the idea that LITE-1 is a low-affinity diacetyl receptor that drives avoidance or the Ca2+ responses in ASK, since it is still present in lite-1 mutants. The authors then use a strain that expresses LITE-1 in the body wall muscles and show this expression is sufficient to engender them with sensitivity to diacetyl, as measured through altered swimming and hypercontractility. The authors interpret this result as LITE-1 may act as a diacetyl receptor. The authors test whether a structurally similar molecule, 2,3-pentanedione, shows similar effects, and they find it does. Alpha-fold modeling and molecular docking analysis show where diacetyl might bind to the LITE-1 protein. They then test whether lite-1 mutants show chemotaxis defects to other molecules, as seen with diacetyl. Generally, they find that the observed diacetyl responses are unique, although lite-1 mutants do lose their avoidance response to 2,3-pentanedione. However, unlike the acquisition of diacetyl attraction in lite-1 mutants, 2,3 pentanedione avoidance is *lost*; it is not switched to attraction. Overall, I felt the description of the results and their implications could have been more in-depth. Further, the evidence that LITE-1 is a chemoreceptor itself, rather than acting in some way to shape chemoreceptor responses (via light or otherwise), remains unclear, as conceded by the authors.

Strengths:

Overall, the study follows up on an interesting and useful result. The experiments as presented are generally well-conceived and performed. The authors use a variety of behavioral and imaging approaches to test how LITE-1 mediates diacetyl avoidance.

Weaknesses:

The study is missing experiments needed to resolve whether LITE-1 is doing what they propose. The evidence that LITE-1 is a diacetyl receptor is lacking support since lite-1 mutants have their avoidance and calcium responses flipped, which would not be expected if it were acting solely as an avoidance receptor. Presumably, the authors are concluding that the attractive response that is left in the lite-1 mutant is mediated by ODR-10, but that experiment is not shown.

We interpret the shift from avoidance to attraction in lite-1 mutants as consistent with the loss of an aversive sensory component in the presence of an underlying attractive response to diacetyl. We initially hypothesised that this residual attraction was mediated predominantly by ODR-10. To test this, we now generated and analysed lite-1; odr-10 double mutants. The double mutants retained an attractive response to diacetyl, indicating that ODR-10 alone does not account for the attraction observed in the absence of LITE-1 and that additional receptors or sensory pathways are likely to contribute. This finding is consistent with previous studies where loss of ODR-10 did not lead to a complete loss of diacetyl responsiveness.

Similarly, the authors concede that "the use of lite-1 point mutants that affect specific LITE-1 function, such as light sensing, channel gating, or binding pocket, could further elucidate LITE-1 mechanisms." This reviewer agrees, and such experiments designed to localize diacetyl binding site(s) would be necessary to conclude definitively that LITE-1 is a diacetyl receptor. The body wall muscle assay used or some other heterologous experimental system could work for such a structure-function analysis. A concern is whether the extensive number of LITE-1 point mutants described in the literature affect cell surface expression vs. receptor function, which might complicate the interpretation of a result showing loss of diacetyl responses.

We agree that structure-function analysis using LITE-1 point mutants could help identify regions or residues that contribute to the diacetyl response and is an important future direction for research, which we have included in the discussion.

Reviewer #2 (Public review):

Summary:

Koh and colleagues investigate the broader sensory role of LITE-1, a gustatory receptor previously linked to UV light detection in C. elegans. Their study explores whether LITE-1 also mediates avoidance of specific chemical stimuli-namely, high concentrations of diacetyl and 2,3-pentanedione. They show that LITE-1 is required in the ADL and ASK neurons for calcium responses to diacetyl, and that its expression in body-wall muscles is sufficient to trigger hypercontraction upon odorant exposure. Molecular docking suggests both odorants may directly bind to LITE-1 with micromolar affinity. These findings suggest LITE-1 may act as a multimodal receptor for both light and chemical stimuli.

Strengths:

(1) Methodological Precision: The study is technically strong, with well-executed calcium imaging and quantitative behavioral assays that clearly show neural and muscular responses to chemical stimuli.

(2) Novelty and Scope: The work presents a compelling case for LITE-1 functioning as a multimodal sensor, which is an intriguing expansion of its known role.

(3) Potential Impact: If validated, the findings could significantly advance the understanding of sensory integration in C. elegans, and the tools developed may be broadly useful to the research community.

(4) Relevance to the Field: The study adds to evidence that C. elegans uses non-canonical sensory pathways and may inspire further exploration of multimodal receptor functions in other systems.

Weaknesses:

(1) Lack of Rescue Experiments: The absence of rescue experiments makes it difficult to definitively link the observed phenotypes to loss of lite-1.

We have now performed the rescue experiment expressing lite-1 in ADL, and showed that LITE-1 in ADL is sufficient for avoidance, although it is not a complete rescue to wild-type levels.

(2) Single Loss-of-Function Approach: The reliance on a single genetic mutant limits interpretability. Additional strategies such as RNAi (e.g., neuron-specific knockdown) would provide stronger evidence.

We observed the loss of avoidance in three independent lite-1 alleles. Combined with the new cell-specific rescue experiment, we think this provides sufficient support for the conclusion that the phenotype is due to loss of lite-1 function.

(3) Unclear Neuronal Contribution: While calcium responses in ADL and ASK are reduced, it's unclear which neuron(s) are necessary for behavioral avoidance. Cell-specific rescue or knockdown experiments are needed.

We have expressed lite-1 genomic DNA under the ADL-specific promoter srh-220, which restored the avoidance phenotype, although it is not a complete rescue of wild-type behaviour. Together with calcium imaging data, this suggests that proper avoidance likely requires input from both ADL and ASK neurons.

(4) Unvalidated Docking Data: The molecular docking predictions lack experimental validation. Site-directed mutagenesis would be needed to support claims of direct interaction.

We agree that the docking data does not in itself establish direct binding (we think the muscle expression and paralysis provides stronger evidence). Based on previously reported docking experiments, we wanted to check if diacetyl could occupy the same binding pocket. We have now also included docking data of the other odorants from the chemotaxis assays in the manuscript.

(5) Limited Odorant Specificity Testing: Docking analysis does not include non-binding odorants, making it difficult to assess binding specificity.

We agree and have now included docking data of the other odorants from the chemotaxis assays. 2-butanone, which is avoided by lite-1 mutants, was predicted to have a slightly higher binding affinity for LITE-1 than 2,3-pentanedione. This highlights the need to interpret the in silico docking data together with real experimental data, rather than using the computational predictions alone to infer functional receptor activation.

(6) Incomplete Quantification: Some calcium imaging results (e.g., in AWA neurons of unc-13 mutants) lack statistical comparisons, which limits their interpretive value.

We have generated the scatter plots of calcium imaging responses across the different sensory neurons, and the statistical significance was assessed using two-sided t-tests with FDR correction, which is now included in the manuscript.

Reviewer #3 (Public review):

In this work, Brown and colleagues report that the photosensor protein LITE-1 of the nematode C. elegans may also be a chemosensor that can be activated by high concentrations of the compound diacetyl. LITE-1 was described as a putative ion channel of the gustatory receptor family, which is mainly constituted by insect odorant receptors. These form tetrameric ion channels that can be activated by odorants. Specificity is achieved by forming heteromeric channels from three copies of the odorant receptor co-receptor (ORCO) and another subunit that resembles ORCO in the pore-forming C-terminus, but brings in a binding site for the respective odorant. LITE-1 has a very similar structure, according to Alphafold3 predictions, and also carries a binding pocket. In LITE-1, this was proposed to be occupied by a light-absorbing molecule that activates the channel when a photon is absorbed. Alternatively, compounds generated by absorption of high-energy photons may be formed in vivo and bound by the LITE-1 binding pocket. Koh et al. now demonstrate that another, non-light-activated compound, diacetyl, at high concentrations, can activate cells expressing LITE-1. Such (chemosensory) cells are also responsible for the avoidance of high concentrations of diacetyl. LITE-1 activation in excitable cells, i.e, muscles, causes strong body contraction and paralysis, and the authors show that this is also the case when diacetyl is presented. The authors further present molecular docking studies showing that diacetyl could occupy the binding pocket of LITE-1. Last, they show that another compound chemically resembling diacetyl, i.e., 2,3-pentanedione, can also induce avoidance in a LITE-1 dependent manner, though not as potently.

The data are intriguing, and the demonstration of LITE-1 being a diacetyl chemosensor is interesting. Yet, there are a few questions arising that the authors should address.

The authors identified mutants lacking diacetyl responses. In their chemotaxis assay (Figures 1A, B), they show that lite-1 mutants do not avoid high concentrations of diacetyl. However, the animals actually showed attraction, as the chemotaxis index was positive. If the lite-1 animals were insensitive, they should be indifferent, and the chemotaxis index should be close to zero. This means, other neurons contribute to the diacetyl response, and the result of these neurons being activated means/remains attraction? If so, the authors need to rule out any effects of these neurons on the effects they attribute to LITE-1 in the other assays.

We have tested tax-4 mutants in the chemotaxis assay and found that, contrary to the predicted chemotaxis index of zero, these animals retained strong avoidance of high concentrations of diacetyl. This indicates that tax-4 mutants are not chemosensory null for this stimulus and that TAX-4 independent sensory pathways contribute to high diacetyl avoidance. We agree that these experiments cannot completely rule out indirect neuronal effects. We have therefore revised the text to acknowledge this limitation. Nevertheless, the rapid paralysis and contraction observed when LITE-1 is expressed specifically in body-wall muscle in a lite-1 mutant background support the idea that LITE-1 is sufficient to confer a diacetyl-evoked response in these cells.

The effect of diacetyl on muscle cells (Figure 3C) is pretty rapid, i.e., already during 1 minute after application, the animals are almost maximally contracted. How fast is it really? Can the authors provide a time course with more time points during the first minute? This is a relevant question, as the compound would have to either pass the worm cuticle or enter through the gut and diffuse through the body to reach the muscle cells. Can one expect this to occur within (less than) a minute? In this context, the authors need to rule out that other mechanisms may be at play. E.g., diacetyl may be immediately sensed by ciliated chemosensory neurons that might release a signaling molecule that leads to activation of LITE-1 in muscles, or that sensitizes it somehow, responding to light used for filming animals. The authors should repeat this assay in a lite-1 mutant background.

We repeated the paralysis assays under red-filtered illumination to minimise potential effects of light, with animals maintained in darkness from hatching to adulthood. We also included lite-1 mutants to assess whether neuronal LITE-1 contributed to the paralysis response. In addition, the assay was repeated with more frequent time points, revealing that paralysis and body contraction occurred within 10 s and neuronal LITE-1 does not contribute to the effect.

Furthermore, the authors tested unc-13 mutants to rule out indirect effects on the neurons recorded. Likewise, they should eliminate neuropeptide signaling via unc-31 mutants (a recent paper cited by the authors showed involvement of neuropeptide signaling in LITE-1-mediated light avoidance behavior).

We agreed and have acknowledged and discuss in the manuscript that contributions from gap junction-mediated communication, neuropeptide signalling and other chemosensory pathways cannot be excluded.

Last, to demonstrate that effects are not indirect in response to chemosensory neurons, the authors should repeat the contraction or swimming assay in a tax-4 mutant, which largely lacks chemosensation. This also applies to the chemotaxis assay. Animals should exhibit a chemotaxis index to diacetyl of zero, then.

We have tested tax-4 mutants, and like wild-type animals, retained strong avoidance of high concentrations of diacetyl, indicating that TAX-4-independent sensory pathways contribute to this response. This indicates that tax-4 mutants are not chemosensory null for this stimulus and that TAX-4-independent sensory pathways contribute to high diacetyl avoidance. Therefore, repeating the contraction or swimming assay in a tax-4 background would not completely exclude indirect input from other chemosensory neurons. In addition, rapid paralysis and contraction were observed when LITE-1 is expressed specifically in body-wall muscle in a lite-1 mutant background, and together with the calcium imaging and rescue data, they support a role for ADL and ASK in mediating high diacetyl avoidance. The tax-4 chemotaxis data is now included in the manuscript.

Does diacetyl activate other neurons expressing LITE-1? A number of cells express LITE-1 at high levels, which the authors have not tested (they restricted their analyses to chemosensory neurons). This is important to address because it leaves the possibility that LITE-1 requires a specific partner only present in these chemosensory neurons to detect diacetyl. This partner would have to be present also in muscles, where diacetyl could activate ectopically expressed LITE-1. According to CeNGEN scRNAseq data, cells expressing LITE-1 can be identified. The ADL and ASH neurons actually come up only at the lowest threshold, so some of the other cells showing much higher levels of LITE-1 mRNAs, i.e., AVG, ALM, PLM, ASG, PHA, PHB, AVM, RIF, or some pharyngeal neurons, should be tested. ASG was among the cells the authors recorded from, but this neuron did not show a response.

We have acknowledged and discuss in the manuscript that other non-sensory neurons may contribute to the avoidance behavioural, and which should be the future direction for investigation.

The authors need to show that diacetyl responses of ADL and/or ASK can be rescued by expressing LITE-1 specifically in these neurons in a lite-1 mutant background.

We have expressed lite-1 genomic DNA under the ADL-specific promoter srh-220, which restored the avoidance phenotype, although it is not a complete rescue of wild-type behaviour. Together with calcium imaging data, this suggests that proper avoidance likely requires input from both ADL and ASK neurons.

Molecular docking studies are not described in detail. How was this done?

Molecular docking was performed in two stages. First, diacetyl was docked to the tetrameric LITE-1 model using DynamicBind without a predefined binding pocket. The generated complexes were ranked using the DynamicBind confidence score, and the highest ranked poses were used to identify the candidate binding site. The top DynamicBind pose was then used to define the box region for redocking with Gnina. Gnina poses were ranked using the CNN score. A more detailed molecular docking procedure has now been updated in the Methods section.

Diacetyl is a very small molecule. How well can docking algorithms assess this at all?

We agree that the small size of diacetyl limits the precision of docking scores because it forms relatively few protein contacts. However, its small size and limited conformational flexibility also simplify pose sampling. To increase robustness, we used two conceptually different docking approaches. First, DynamicBind was used without a predefined binding pocket to identify candidate binding regions while allowing ligand-associated protein conformational adjustments. Second, the resulting pocket was subjected to focused redocking and CNN-based pose ranking with Gnina. The results are interpreted as a structural hypothesis for the probable binding site and relative affinity ranking, rather than as definitive proof of binding or an accurate quantitative affinity measurement.

Did the authors preselect the binding pocket, or did the algorithm sample the entire molecular surface of the LITE-1 model and end up with the binding pocket?

The binding pocket was not predefined. Diacetyl was first docked to the tetrameric LITE-1 model using DynamicBind without specifying pocket residues or grid coordinates. DynamicBind therefore performed global, pocket-agnostic docking. The highest-ranked poses identified a candidate pocket, which was then used for focused redocking with Gnina.

The latter would be very convincing. The authors should provide control docking experiments with other molecules that caused avoidance in their hands (i.e. benzaldehyde, 2,4,5,trimethlythiazole, isoamyl alcohol, nonanone, octanone), but did not activate LITE-1. Also, they should try docking molecules related to diacetyl, and if there are some that do not dock under the same conditions, such molecules should be used in a behavioral experiment. Ideally, they should also not activate LITE-1. Examples could be, e.g., diacetyl monoxime or 2,4-pentanedione.

We have now included docking data of the other odorants from the chemotaxis assays. 2-butanone, which is avoided by lite-1 mutants, was predicted to have a slightly higher binding affinity for LITE-1 than 2,3-pentanedione. This highlights the need to interpret the in silico docking data together with real experimental data, rather than using the computational predictions alone to infer functional receptor activation.

Last, the authors should provide a PDB file with the docked diacetyl to allow readers to assess the binding for themselves. Since a large number of mutations of LITE-1 have been reported, it may be that amino acids shown to be essential for LITE-1 function are also required for diacetyl binding. If so, this could be backed up with an experiment.

We agree that structure-function analysis using LITE-1 point mutants could help identify regions or residues that contribute to the diacetyl response and which we have highlighted in the discussion as an important future direction for research. Additionally, we have now provided the PDB file containing a representative DynamicBind derived docking pose of diacetyl within the LITE-1 binding pocket.

Recommendations for the authors:

Reviewer #1 (Recommendations for the authors):

(1) lite-1 mutant animals, as described, fail to avoid' high concentrations of diacetyl, but isn't it more accurate to say that the valence of the response is changed from avoidance to attraction? Do the authors believe that attraction is mediated by ODR-10? Can you build an odr-10; lite-1 double mutant and determine if they lose this attraction to high concentrations of diacetyl and/or 2,3 pentanedione?

Our initial interpretation was that, in the absence of LITE-1-mediated avoidance, attraction to high concentrations of diacetyl is driven by the low-concentration receptor ODR-10. Interestingly, however, the lite-1; odr-10 double mutants remained strongly attracted to high concentrations of diacetyl, suggesting that this phenotype is independent of ODR-10 and may instead be mediated by other, less specific odorant receptors. This is consistence with the odr-10 mutants not completely losing attraction to low concentration of diacetyl, suggesting the involvement of other potential/putative receptors (Sengupta et al., 1996; Taniguchi et al., 2015). The lite-1; odr-10 double mutant data is now added in the results section (lines: 82 to 90; Figure S1B).

(2) For ADL, yes, it seems like lite-1 mutants have a reduced diacetyl response, but the ASK response seems... different. While it goes up (slowly) in wild-type, cellular Ca2+ levels in ASK (and maybe ADL) are *reduced* by diacetyl in lite-1 mutants. Can the authors comment on this, and the behavioral responses change in valence?

ADL and ASK are involved in both attractive and aversive responses so it is possible that the attractive component of the diacetyl response suppresses their activity. In the absence of LITE-1 activation, the observed decrease in calcium responses may reflect the unopposed inhibitory input. The slower decay of the calcium signal in ASK neurons of unc-13 mutants further supports the presence of additional inhibitory signals influencing their activity. This explanation is now included in the results section (lines: 101 to 117).

(3) ADL and ASK calcium traces in unc-13 mutants look generally similar and lack the effects seen in lite-1 mutants. Does that mean the lite-1 effect is in cells other than ADL or ASK? Can the authors spend more time discussing these differences?

We acknowledge that LITE-1 is expressed in multiple cell types beyond chemosensory neurons, and that non-chemosensory neurons may also contribute to the observed phenotype. In the previous version, we had highlighted the interneuron AVG as a potential contributor, given its role in light-induced escape. In the revised discussion, we have now expanded this section to include additional possible contributors such as the LITE-1-expressing phasmid neuron PHA and the pharyngeal interneurons I2, both of which have been implicated in hydrogen peroxide sensing (lines: 177 to 181).

(4) Chemotaxis responses to diacetyl and 2,3-pentanedione in lite-1 mutants are rather different. Diacetyl switches from repulsive (CI < 0) to *attractive* (CI > 0), which is not what would be expected for mutations that eliminate a receptor. In contrast, the 2,3-butanedione responses are more what would be predicted: diacetyl goes from inhibitor to no effect (CI ~0). Again, if the authors feel that this is because of the ODR-10 function, can they discuss whether 2,3-butanedione is predicted to bind ODR-10 like diacetyl?

We think a switch to attraction is expected for the removal of a receptor for an aversive signal, in an attractive background signal. We did indeed think that this attraction was mediated by odr-10, but the double mutant results now show other receptors must be involved. This is also not wholly unexpected since previous work has identified other receptors of high-concentration diacetyl and the original odr-10 paper didn’t report a complete absence of diacetyl response, suggesting the presence of other receptors that mediate attraction to diacetyl (Sengupta et al., 1996; Taniguchi et al., 2014). The new data is now added in the results section (lines: 82 to 90; Figure S1B; Supplementary video 1 and 2).

(5) Were the behavior experiments performed in the dark? I realize the calcium imaging experiments and some of the video behavior recordings are not possible in complete darkness, but maybe the authors made efforts to exclude visible light effects (e.g., infrared illumination, etc.) in some assays that might help determine whether light plays *no* role in the effects observed. Alternatively, the authors could try repeating their chemotaxis experiments in the dark or at least communicate in the methods that this was not viewed as a concern (and why). As the authors propose and discuss LITE-1 modulating diacetyl responses via light sensation as a possibility, it is incumbent upon them to communicate the steps they took to overcome this concern for themselves.

We acknowledge this and have performed a chemotaxis experiment to compare assays performed under dark and ambient light conditions, and no significant differences were observed (results section: lines 75 to 80; Figure S1A, material and methods section: 241 to 246). Therefore, subsequent chemotaxis assays were carried out under ambient light while avoiding exposure to strong illumination.

Paralysis assays were repeated under red-filtered illumination to minimise light effects, with animals maintained in darkness from hatching to adulthood. Additionally, the assay was expanded to include lite-1 mutants, ruling out contributions from neuronal LITE-1 to paralysis. The new data is now incorporated into the result section (diacetyl; lines: 135 to 137, figures 4 and S3; 2,3-pentanedione; lines: 152 to 154, figures 4 and S6), materials and methods section have been updated to reflect these changes (lines: 250 to 253 and 257 to 260).

Reviewer #2 (Recommendations for the authors):

Minor Issues:

(1) Pmyo-3::LITE-1 worms shrink in the absence of odorants (Figures 3C, 4D); possible effects of ambient light should be discussed.

We acknowledge the possibility that worms expressing LITE-1 in body-wall muscle experience minor contractions under ambient light, though it is not sufficient to cause paralysis.

To minimise potential light-induced effects, the paralysis assays were repeated with red-filtered illumination to reduce light stimulation of LITE-1. Animals were maintained in darkness from hatching to adulthood, and in the updated assay, worm length remained relatively constant.

The new data (Figures 3C, 4E, S4C and S6C) and materials and methods section has been updated accordingly (lines: 250 to 253 and 257 to 260).

(2) The title is misleading, as ASH does not show altered activity in lite-1 mutants and should be removed from the claim.

ASH has been removed from the title (line: 97).

(3) Specific Kd values should be provided for the reported micromolar binding affinities.

The values from DynamicBind and Gnina are provided in Figure S5.

Recommendations:

(1) LITE-1, a member of the gustatory receptor family, was previously shown to mediate UV light responses in C. elegans. In this study, Koh and colleagues demonstrate that LITE-1 is also required for the nematode's avoidance of high concentrations of diacetyl - an odorant that is attractive at low levels but aversive at higher concentrations. Using calcium imaging, the authors show that LITE-1 is necessary in the sensory neurons ADL and ASK for calcium transients in response to high concentrations of diacetyl. Additionally, they find that expressing LITE-1 in body-wall muscles causes hypercontraction upon diacetyl exposure. Similar LITE-1-dependent responses were observed for 2,3-pentanedione, another structurally related odorant. Molecular docking analyses suggest that both diacetyl and 2,3-pentanedione directly bind to LITE-1 with micromolar affinity.

These findings are intriguing and have the potential to significantly advance our understanding of LITE-1 as a multimodal sensory receptor. However, several major issues need to be addressed to support the authors' conclusions:

(1) Rescue experiments are missing. The authors should rescue at least one lite-1 mutant to confirm that the observed avoidance defects are specifically due to loss of lite-1.

Because the avoidance defect was observed in three independent lite-1 alleles, we think background mutations are unlikely to be causal. We have now also performed a rescue experiment with lite-1 expressed in ADL neurons. In this strain, attraction is restored. The new data have been incorporated into the results section (lines: 119 to 121; Fig. 2D).

(2) Alternative loss-of-function approach. To strengthen the findings, the authors should use a different method to disrupt lite-1 function-such as RNAi by feeding or cell-specific RNAi driven by the lite-1 promoter (see PMID: 17459615).

We believe the multiple alleles (Fig. 1B) and new cell-specific rescue experiment (Fig. 2D) provide sufficient support for the conclusion that the phenotype is due to loss of lite-1 function.

(3) Clarify the role of ADL and ASK neurons. While calcium imaging data show reduced activity in these neurons in lite-1 mutants, it remains unclear whether lite-1 is required in ADL, ASK, or both for avoidance behavior. Cell-specific rescue or RNAi experiments, along with additional calcium imaging, are needed to determine the contribution of each neuron.

We agree that more in-depth work will be required to dissect the neuronal pathways involved in LITE-1-mediated diacetyl avoidance. We have avoided making specific comments on how exactly the observed imaging results relate to the behavioural phenotype. The new rescue experiment expressing lite-1 in ADL does at least show that LITE-1 in ADL is sufficient for avoidance, although it’s not a complete rescue to wild-type levels (lines: 119 to 121; Fig. 2D).

(4) Validation of molecular docking results. While molecular docking suggests direct binding of odorants to LITE-1, experimental validation is needed. Mutations that reduce predicted binding affinity (engineered in transgenes or via CRISPR) should be tested for functional impact on avoidance behavior.

We agree the docking does not in itself establish direct binding (we think the muscle expression and paralysis provides much stronger evidence). Based on previously reported docking experiments, we were simply curious whether diacetyl would be predicted to occupy the same binding pocket. We have now updated the discussion in the use of lite-1 mutants to test for impact on diacetyl avoidance (lines: 188 to 191).

(5) Include analysis of non-binding odorants. Docking results should also be presented for odorants that did not elicit LITE-1-dependent avoidance, to help establish specificity.

We have now included docking data of odorants from the chemotaxis assays, with the corresponding docking values shown in Supplementary Figure 5. 2-butanone, which is avoided by lite-1 mutants, was predicted to have a slightly higher binding affinity for LITE-1 than 2,3-pentanedione. This highlights the need to interpret the in silico docking data together with real experimental data, rather than using the computational predictions alone to infer functional receptor activation.

(6) Figure 2C concerns. In neurons such as AWA, calcium transients in unc-13 mutants appear reduced compared to wild-type. A statistical comparison for all the neurons should be included to assess significance.

Scatter plots of calcium imaging responses across the different sensory neurons were generated, and statistical significance was assessed using two-sided t-tests with FDR correction. Only ADL and ASK neurons showed significant differences between lite-1 mutants and wild-type animals. No significant differences were observed in any neuronal pairs between unc-13 mutants and wild-type, including AWA neurons, although the difference is close to significant (p = 0.07). The scatter plots were now included as Supplementary Figure 3.

Minor points:

(1) Figures 3C and 4D: Pmyo-3::LITE-1 worms appear to shrink even without diacetyl or 2,3-pentanedione. Could this be due to ambient light? The authors should discuss this possibility.

We acknowledge the possibility that worms expressing LITE-1 in body-wall muscle experience minor contractions under ambient light, though it is not sufficient to cause paralysis. We repeated the paralysis assays with red-filtered illumination to reduce light stimulation of LITE-1. Animals were maintained in darkness from hatching to adulthood, to minimise potential light-induced effects, and in the updated assay, worm length remained relatively constant.

The new data (Figures 3C, 4E, S4C and S6C) and materials and methods section has been updated accordingly (lines: 250 to 253 and 257 to 260).

(2) Title revision needed: The title "Chemosensory neurons ADL, ASK, and ASH are involved in avoidance of diacetyl" is misleading, as calcium transients in ASH appear unaffected in lite-1 mutants. The title should reflect the actual data.

ASH have been removed from the title (line: 97).

(3) Binding affinity clarification: The authors report micromolar binding affinity for LITE-1 but should provide specific dissociation constants for clarity and completeness.

The values from DynamicBind and Gnina are now provided in Supplementary Figure 5.

Reviewer #3 (Recommendations for the authors):

How did the authors measure body length if the animals were swimming in the diacetyl solution? Standard 6-well plates have an area of roughly 10 cm², meaning that if one adds 1 ml, the liquid level should be 1 mm. The animals would be able to move in 3D, so it is likely that animals swim up and down and do not move in one flat plane, i.e., head and tail would be out of focus, and only a projection image would be recorded that would lead to an underestimation of actual worm length.

We acknowledge that this issue may led to an underestimation of worm length in the previous assay. However, every effort was made to exclude worms that moved out of focus. The paralysis assay has since been modified to include spreading a thin layer of solution across the worms, which keeps them mostly in focus, particularly those that are paralysed. Worms that were partially out of focus were excluded from the analysis.

We have incorporated the new data into the results section, reflected in the updated Figures 3C, 4E, S4C, and S6C. Corresponding revisions have also been made in the materials and methods (lines: 250 to 253 and 257 to 260).

Could diacetyl be a compound that results from UV absorption in cells? This may be worth discussing. What could be the precursor molecule?

We are not aware of such a precursor, but we cannot rule it out. Even if UV absorption in cells leads to diacetyl production, it is likely that the resulting diacetyl levels are insufficient to activate LITE-1, as our data suggest that LITE-1 functions as a receptor for high concentrations of diacetyl. We have updated the discussion accordingly (lines: 169 to 173).

In lines 57-61, the references to Edwards 2008 and Ward 2008 do not seem to fit the statements made in this sentence.

The inclusion of Edwards et al., 2008 was an error, and it has now been removed. Ward et al., 2008 demonstrated that ASJ phototransduction requires cGMP and CNG channels, stating that “Our studies indicate that C. elegans photoreceptor cells also employ CNG channels and the second messenger cGMP for phototransduction.”

Version 2

Abstract

The nematode C. elegans does not have eyes but can respond to aversive UV and blue light stimulation and even distinguish colours. The gustatory receptor homolog LITE-1 was identified in forward genetic screens for worms that failed to respond to blue light stimulation. When LITE-1 is expressed in body-wall muscles, it causes contraction in response to blue light suggesting that LITE-1 is both necessary and sufficient for blue light response. Here we show that in addition to light avoidance, LITE-1 is also required for worms’ avoidance of high concentrations of diacetyl, an odorant that is attractive at low concentrations. Like blue light, diacetyl causes muscle contraction in transgenic worms engineered to express LITE-1 in body-wall muscles. These data are consistent with a direct chemoreceptor function for LITE-1 which would make it a multimodal sensor of aversive stimuli.

Introduction

Sensory perception is fundamental to how organisms interact with their surroundings. Animals, including Caenorhabditis elegans, detect and respond to a wide range of environmental cues such as light and odorants. Despite having a compact nervous system of only 302 neurons, C. elegans exhibits sensory capabilities that allow it to effectively navigate and adapt to its environment (Bargmann, 2006; Cook et al., 2019). Odorants such as volatile organic compounds are detected via specialised olfactory receptor neurons. These neurons express specific receptors that bind to specific molecules and trigger distinct behavioural responses (Bargmann, 2006; Bargmann et al., 1993; Bargmann & Horvitz, 1991). Diacetyl, a volatile organic α-diketone found in C. elegans’ natural environment elicits opposing behaviours depending on its concentration. At low concentrations, diacetyl is sensed by the olfactory receptor ODR-10, promoting attraction (Itskovits et al., 2018; Sengupta et al., 1996; Zhang et al., 1997). In contrast, higher concentrations induce an avoidance response, thought to be mediated by the receptor SRI-14 (Taniguchi et al., 2014).

In addition to chemical cues, C. elegans also detects and responds to light, despite lacking eyes or classical photoreceptors. This light sensitivity is mediated by LITE-1, a member of the Gustatory Receptor family (Montell, 2009). LITE-1 was first identified through genetic screens for mutants defective in light avoidance behaviour (Edwards et al., 2008). Unlike canonical photoreceptors such as rhodopsins or opsins, LITE-1 shares little sequence similarity with known light sensing proteins (Edwards et al., 2008; Gong et al., 2016; Liu et al., 2010). Further underscoring its uncanonical nature, LITE-1 adopts an inverted membrane topology that is unlike classical G protein-coupled receptors (GPCRs) (Gong et al., 2016).

LITE-1 phototaxis behaviour is driven by multiple light responsive sensory neurons, including ASJ, ASH and ASK (Liu et al., 2010; Ward et al., 2008; Zhang et al., 2022). Signal transduction occurs via G protein signalling, upregulation of cGMP levels and activation of cyclic nucleotide-gated (CNG) channels, leading to a rapid avoidance response (Edwards et al., 2008; Liu et al., 2010; Ward et al., 2008). Recent studies suggest that LITE-1 may function as a light activated ion channel with a putative ligand binding pocket that is activated by ultraviolet and blue light via key tryptophan residues (Edwards et al., 2008; Gong et al., 2016; Hanson, Scholüke, et al., 2023; Ward et al., 2008).

Here we show that beyond its role in photoreception, LITE-1 is also required for the avoidance of high concentrations of diacetyl and that diacetyl induces muscle contraction in transgenic worms expressing LITE-1 in body-wall muscles. Our findings support the notion that LITE-1 functions as a multifunctional receptor involved in both light and chemical sensing.

Results

LITE-1 is required for diacetyl and blue light avoidance

A recent screen of a panel of worm strains that collectively contained knockouts of all non-essential GPCRs identified a triple mutant that failed to avoid high concentrations of diacetyl (Pu et al., 2023). One of the mutated genes was lite-1. We performed a chemotaxis assay to test single mutants of several loss-of-function alleles of lite-1 and found they all failed to avoid high concentrations of diacetyl (Fig. 1A and 1B) while remaining attracted to lower concentrations (Fig. 1C). As LITE-1 is sensitive to light, we next examined whether ambient light affected these responses by comparing chemotaxis under light and dark conditions. Under both conditions, wild-type animals strongly avoided high concentration of diacetyl, whereas lite-1 mutants still failed to do so, indicating that this phenotype is light independent (Fig. S1A). As expected, these lite-1 mutants also no longer respond to blue light stimulation (Fig. 1D) (Edwards et al., 2008).

(A) Schematic of chemotaxis assay setup. (B) Chemotaxis index of wild-type animals and mutants with different alleles of lite-1 that have defects in diacetyl response. P-values are from unpaired two-tailed t-tests, n = 4 biological replicates, each an average of 4 plates. (C) Chemotaxis index of wild-type and lite-1 (yum2880) animals with different diacetyl concentrations (5 μl). P-values are from unpaired two-tailed t-tests, n = 3 biological replicates, each an average of 4 plates. (D) Time series of wild-type (blue line) animals and mutants with different alleles of lite-1 (orange line) when stimulated with blue light (blue shaded box). n = 3 biological replicates, each an average of 12 wells.

We next examined whether the lite-1 mutants’ attraction to high concentrations of diacetyl depends on the low-concentration diacetyl receptor ODR-10 by analysing lite-1; odr-10 double mutants. The double mutants remained attracted to high concentrations of diacetyl (Fig. S1B). This attraction may result from the activation of other, less specific odorant receptors whose signals become unmasked in the absence of LITE-1 mediated inhibition. This is consistent with earlier reports where worms do not completely lose attraction to low concentrations of diacetyl in the absence of ODR-10, suggesting the presence of additional low concentration receptors (Sengupta et al., 1996; Taniguchi et al., 2014). We also examined whether the avoidance behaviour depends on TAX-4 mediated chemosensory signalling. Like wild-type worms, tax-4 mutants retained strong avoidance to high concentration of diacetyl, suggesting that TAX-4 independent pathways contribute to the avoidance response (Fig. S2). Consistence with additional aversive pathways, observations of lite-1 mutants show that they do not reach the highest concentration of diacetyl (Supplementary video. 1 and 2).

Chemosensory neurons ADL and ASK are involved in avoidance of diacetyl

To identify neurons involved in the response to high concentrations of diacetyl, we used calcium imaging to record 11 pairs of sensory neurons in worms exposed to a pulse of diacetyl (Fig. 2A). In lite-1 mutants, the LITE-1-expressing sensory neurons ADL and ASK showed a decrease in calcium levels (Fig. 2B and S3A). Since ADL and ASK sense both attractive and aversive stimuli, it is plausible that the attractive response to diacetyl suppresses their activity. In the absence of LITE-1-mediated activation, the observed decrease in calcium responses may reflect the unopposed inhibitory signal input. In contrast, another LITE-1 expressing neuron, ASH, showed a normal response compared to wild-type animals, which is likely due to its expression of SRI-14 (Fig. 2B and S3A). SRI-14 has previously been shown to respond weakly to high concentration of diacetyl (Taniguchi et al., 2014). The response of the other tested neurons was not significantly different between lite-1 mutants and wild-type animals (Fig. 2B and S3A). Additionally, we observed that the unc-13 mutants, which are impaired in synaptic activity, had little effect on ADL response, but had a slower decay of the calcium signal in ASK neurons compared to wild-type animals (Fig. 2C and S3B) (Richmond et al., 1999). This supports the presence of additional inhibitory signals influencing ASK activity. Together, these data support the involvement of these neurons, although contributions neural signalling through gap junctions or via neuromodulation by other receptors could not be ruled out.

(A) Confocal image of labelled sensory neurons and calcium imaging traces showing the responses of three neurons when exposed to high concentrations of diacetyl. (B) Neuronal activity of wild-type controls with lite-1 mutants. (C) Neuronal activity of wild-type controls to unc-13 mutants. Grey shaded box denotes the presence of diacetyl. Shaded area around the activity dynamics denotes standard error. (D) Chemotaxis index of wild-type animals and rescue mutants expressing lite-1 cDNA from the srh-220 promoter in ADL neuron. P-values are from unpaired two-tailed t-tests, n = 4 biological replicates, each an average of 4 plates.

We next expressed lite-1 genomic DNA under the ADL specific promoter srh-220, which restored the avoidance phenotype, although it is not a complete rescue of wild-type behaviour (Fig. 2D) (Hao et al., 2010). Together with calcium imaging data, this suggests that proper avoidance likely requires input from both ADL and ASK neurons.

LITE-1 is sufficient for diacetyl response in other cell types

To test whether LITE-1 is sufficient for a response to high diacetyl concentrations, we used a transgenic strain that expresses LITE-1 in body-wall muscles (Edwards et al., 2008). Worms ectopically expressing LITE-1 in body muscle cells evoke muscle contraction and shortening of body length in response to blue light stimulation (Edwards et al., 2008). When these worms are exposed to a 1:10 dilution of diacetyl, they showed a rapid and near complete paralysis whereas wild-type animals continue to swim (Fig. 3A and 3B). These worms also showed significant contraction in average body length, contracting by ∼14.8% after diacetyl exposure (Fig. 3C). Consistent with muscle contraction, these worms also ejected eggs, while wild-type worms do not (Fig. 3D). To rule out contributions of neuronal LITE-1, the assay was repeated in lite-1 (ce314) mutants. In these strains, diacetyl still induced paralysis, muscle contraction, and egg ejection (Fig. S4). These results suggest that LITE-1 is activated by diacetyl. Although it remains possible that it is a diacetyl metabolite that activates LITE-1, this conversion would need to happen within seconds because maximal contraction occurs within 10 seconds of addition.

(A) Percent of worms that are swimming at ten seconds after exposure to buffer untreated or to diacetyl. Wild-type animals continue swimming in buffer and diacetyl (WT, left) whereas worms expressing LITE-1 in body wall muscles (Pmyo-3::LITE-1, right) are mostly paralysed when treated with diacetyl. (B) Sample midline trajectories from wild-type and Pmyo-3::LITE-1 animals showing difference in motion (each trajectory shows 6 seconds of swimming). (C) Worm body length over time after treatment. Wild-type animals do not show a significant length difference in either buffer or diacetyl treatments (left) whereas worms expressing LITE-1 in body-wall muscles show a significant contraction in response to diacetyl (insert showing the average contraction). P-values from paired two-tailed t-test, n ≥ 15 worms. (D) Image of a Pmyo-3::LITE-1 animal that ejected eggs within ten seconds of diacetyl treatment (white arrow). (E) Molecular docking of diacetyl into the AlphaFold-predicted structure of LITE-1 with predicted micromolar affinity.

2,3-pentanedione also interacts with LITE-1

We next tested whether lite-1 is required for avoidance of other high concentration odorants. Of the seven odorants tested, lite-1 mutants showed a defective response to 2,3-pentanedione, a chemical structurally related to diacetyl (Fig. 4A). The effect was weaker than that observed for diacetyl, with worms still avoiding high concentrations of 2,3-pentanedione but more weakly than wild-type animals, independent of ODR-10 (Fig. S1B). Supporting this observation and consistent with a ligand receptor interaction, in silico molecular docking predicts that both diacetyl and 2,3-pentanedione bind to LITE-1 within the putative chromophore binding site, with diacetyl exhibiting the strongest affinity (Fig. 3E, 4B and S5) (Hanson, Scholüke, et al., 2023). Like diacetyl, 2,3-pentanedione also evokes paralysis, muscle contraction and egg ejection of worms expressing LITE-1 in body-wall muscle cells, independent of neuronal LITE-1 (Fig. 4C, 4D, 4E and S6).

(A) Chemotaxis index of wild-type animals and lite-1 mutants for known odorants. P-values are from unpaired two-tailed t-tests, n = 4 biological replicates each an average of 4 plates. (B) Molecular docking of 2,3-pentanedione into the AlphaFold-predicted structure of LITE-1 with predicted micromolar affinity. (C) Image of a Pmyo-3::LITE-1 animal that ejected eggs within ten seconds of 2,3-pentanedione treatment (white arrow). (D) Percent of worms that are swimming at ten seconds after exposure to buffer untreated or to 2,3-pentanedione. Wild-type animals continue swimming in buffer and 2,3-pentanedione (WT, left) whereas worms expressing LITE-1 in body-wall muscles (Pmyo-3::LITE-1, right) are mostly paralysed when treated with 2,3-pentanedione. (E) Worm body length over time after treatment. Wild-type animals do not show a significant length difference in either buffer or 2,3-pentanedione treatments (left) whereas worms expressing LITE-1 in body-wall muscles show a significant contraction in response to 2,3-pentanedione (insert showing the average contraction). P-values from paired two-tailed t-test, n ≥ 15 worms.

Discussion

LITE-1 is best known for its role in light sensation in C. elegans, but how it senses light is not yet clear. In particular, an indirect mechanism in which light produces photoproducts that activate the channel has not been ruled out and a priori has always seemed plausible given that LITE-1 is similar to gustatory receptors that are known to sense chemicals (Montell, 2009). A direct diacetyl response would make a photoproduct-based mechanism more plausible. The identification of a second structurally related compound, 2,3-pentanedione may help in narrowing down photoproducts with similar moieties. Direct and photoproduct light sensing are not mutually exclusive. Based on LITE-1’s predicted structure and its absorption spectrum, it has recently been proposed that UV may be directly sensed while the blue light response may be mediated through a photoproduct or an unidentified chromophore (Hanson et al., 2023). While it is possible that UV absorption within cells may lead to diacetyl production, what this precursor might be remains elusive. Diacetyl generated in this way would need to be produced at a sufficient concentration to activate LITE-1. Based on the behavioural and muscle-contraction data, this occurs at relatively high concentrations.

While our study primarily focused on chemosensory neurons, it remains possible that non-chemosensory neurons expressing LITE-1 may contribute to diacetyl detection or modulating the behavioural response. Indeed, other LITE-1 expressing non-chemosensory neurons, such as the interneuron AVG, which mediates light sensing through the release of the neuropeptide NLP-10, as well as the phasmid neuron PHA and the pharyngeal interneuron I2, which are required for hydrogen peroxide sensing, are possible candidates. (Bhatla & Horvitz, 2015; Dunkel et al., 2025; Quintin et al., 2022). Nevertheless, phototransduction and chemotransduction appear to involve shared sensory neurons, such as ASK and ASH, suggesting a potential overlap in their underlaying signalling mechanisms (Zhang et al., 2022). It is possible that LITE-1 chemosensory response may rely on components necessary for LITE-1 photosensory function, such as G proteins (GOA-1, GPA-3), guanylate cyclase (DAF-11, ODR-1), CNG channels (TAX-2,4) and phosphodiesterase (PDE-1,2,5) (Liu et al., 2010; Ward et al., 2008). Furthermore, the use of lite-1 point mutants that affect specific LITE-1 functions such as light sensing, channel gating, or binding pocket properties could further elucidate LITE-1 mechanisms (Gong et al., 2016; Hanson, Scholüke, et al., 2023).

We have shown that LITE-1 is required for the avoidance of high concentrations of diacetyl and a related compound 2,3-pentanedione. Furthermore, we have shown that LITE-1 is sufficient to make body-wall muscle cells responsive to diacetyl and 2,3-pentanedione, consistent with a direct response. Diacetyl is an ecologically relevant molecule present in C. elegans’ natural environment that has been hypothesised to be involved in prey-finding since low concentrations produced by lactic acid bacteria on rotting fruit is attractive (Choi et al., 2016). Avoidance of high concentrations of diacetyl may therefore also be ecologically relevant. In this case, LITE-1 has likely evolved to be a multimodal receptor of aversive stimuli akin to other multimodal receptors such as TRPV1, which responds to heat, capsaicin, and acidic pH (Zhang et al., 2023).

Material and methods

Bacterial and worm strains maintenance

The E. coil wild-type isolate OP50 was sourced from Caenorhabditis Genetics Center (CGC) and maintained on lysogeny broth (LB) agar under standard laboratory conditions. The following worm strains were used in this study, N2; wild-type isolate (CGC), PR678; tax-4(p678), CHS1173; odr-10(yum2055) (Pu et al., 2023), CHS1610; lite-1(yum2880) (Pu et al., 2023), CHS1721 lite-1(yum2880) odr-10(yum2055) (Pu et al., 2023), KG1180; lite-1(ce314) (Edwards et al., 2008), KG1271; lite-1(ce314) cels37[myo-3prom::lite-1::GFP] (Edwards et al., 2008), KG1272; cels37[myo-3prom::lite-1::GFP] (Edwards et al., 2008), TQ8245; lite-1(xu492) (Zhang et al., 2020), ZAS280; In[osm-6::GCaMP3, osm-6::ceNLS-mCherry-2xSV40NLS] (Iwanir et al., 2019), ZAS371; unc-13(s69) In[osm-6::GCaMP3, osm-6::ceNLS-mCherry-2xSV40NLS] (Bokman, Kalij, et al., 2024; Bokman, Pritz, et al., 2024), ZAS526; lite-1(ce314) In[osm-6::GCaMP3, osm-6::ceNLS-mCherry-2xSV40NLS], were cultured on Nematode Growth Medium seeded with E. coli (OP50) as previously described (Brenner, 1974). Worm synchronization was performed by bleaching unsynchronised gravid adults (Barlow, 2019).

Chemotaxis assay

The chemotaxis assays were conducted under ambient light, but care was taken to prevent exposure to any strong light source at 20°C in 9 cm plates filled with 10 ml of agar (1.7% agar, 1 mM CaCl2, 1 mM MgSO4 and 25 mM K2HPO4 pH 6). Two parallel lines, 1 cm apart, were drawn on the middle of the plates which define the area to place the animals. Two odorant spots were designated on one side of the plate, while two control spots were positioned on the opposite side (Fig. 1A). 1 μl of 1 M NaN3 was added to all four spots to immobilise the animals. Different volumes of each odorant: diacetyl (5 μl), 2,4,5-trimethylthiazole (7.5 μl), benzaldehyde (2.5 μl), isoamyl alcohol (7.5 μl), 2,3-pentanedione (10 μl), 2-nonanone (1 μl), 1-octanol (2.5 μl), 2-butanone (10 μl) and equal volume of ethanol (control) were spotted onto the respective dots. Synchronized day-one adults were washed 3 times in chemotaxis buffer (1 mM CaCl2, 1 mM MgSO4 and 25 mM K2HPO4 pH 6) before being transferred onto the centre of the assay plate where the assay will last for 1 hr at 20 °C. The chemotaxis plates were then stored at 4 °C before counting. The chemotaxis index was calculated as (number of animals on the odorant side – number of animals on the ethanol side) / (sum of all animals on both sides). For chemotaxis conducted in the dark, animals were maintained in darkness from hatching to adulthood. Brief exposure to dim ambient light occurred only during washing and transfer to agar plates, as the latter step had to be performed in a fume hood due to the odorant. Following transfer, plates containing the animals were immediately covered with foil to maintain darkness.

Paralysis assay

The paralysis assays were conducted in the dark (animals were kept in the dark from hatching to adulthood, with ambient light exposure minimised) at 20°C in 9 cm plates filled with 10 ml of agar (1.7% agar, 1 mM CaCl2, 1 mM MgSO4 and 25 mM K2HPO4 pH 6). Synchronized day-one adults were washed 3 times in chemotaxis buffer (1 mM CaCl2, 1 mM MgSO4 and 25 mM K2HPO4 pH 6) before being transferred onto the centre of the plate. Animals were allowed to crawl on the plate for five minutes before a thin layer of ∼20 μl chemotaxis buffer containing either diacetyl or 2,3-pentanedione (1:10) was spread across the worms to completely cover them. Video acquisition began prior to buffer addition, recorded at 8 fps using a Basler acA4024-29um monochrome camera under red-filtered illumination, controlled with pylon Viewer 64-Bit (version 6.2.0.21487). Measurement of animal length was quantified manually using Fiji (Schindelin et al., 2012). Only worms that were in focus were included in the analysis.

Blue light stimulation assay

The blue light stimulation assays were conducted at 20°C in 24-well plates containing agar (1.7% agar, 1 mM CaCl2, 1 mM MgSO4 and 25 mM K2HPO4 pH 6). Synchronized day-one adults were transferred onto the centre of the 24-well plate pre-seeded with OP50. The plates were then transferred onto the multi-camera tracker for 30 min to allow acclimatization to the tracker room in the dark before imaging. Videos were acquired at 25 frames per second using a shutter speed of 25 ms and a resolution of 12.4 μm/pixel. Videos were taken in the following order: 300 s pre-stimulus, blue light stimulus with one 10 s pulse of blue light at 60 s and a 100 s post-stimulus recording. Videos were then processed with Tierpsy Tracker as previously described (Barlow et al., 2022; Javer et al., 2018).

Calcium imaging

Young adult worms were inserted into a microfluidic chip and anesthetized with 10 mM levamisole and allowed to habituate for 10 mins as previously described (Chronis et al., 2007). Recordings began with 1 min of light habituation, after which the stimulus was presented for 1 min The stimulus used was 1:5 diacetyl in CTX buffer (5 mM KH2PO4/K2HPO4 pH 6, 1 mM CaCl2 and 1 mM MgSO4). Recordings were made using a Nikon A1R + confocal laser scanning microscope with a water immersion × 40 (1.15NA) objective controlled by the Nikon NIS-elements software. Z-axis resolution was 0.5-0.8 μm. Acquisition rate was ∼1.5 volumes/second, interpolated to a final framerate of 2 Hz.

Neuron identification and signal extraction

Neuron segmentation and tracking was done on the nuclear mCherry signal using an algorithm developed by (Toyoshima et al., 2016). Neurons were identified visually based on anatomical positions. Neurons that could not be unambiguously identified were removed from analysis. Calcium readings were taken from within 0.9 of the radius of the segmented nuclear sphere. Signal intensity was normalized by baseline activity, defined as the lowest point of a 10-frame running average. All data analyses were performed using in-house MATLAB scripts (Pritz et al., 2023).

Molecular docking

Molecular docking was performed in two stages. Diacetyl was first docked to the tetrameric LITE-1 model using DynamicBind without a predefined binding pocket (Hanson, Scholuke, et al., 2023; Lu et al., 2024). The generated complexes were then ranked using DynamicBind confidence score, and the highest-ranked poses were used to identify the candidate binding site. The top DynamicBind pose was then used to define the box region for redocking with Gnina (McNutt et al., 2021). Gnina poses were ranked using the CNN score. The PDB file containing a representative DynamicBind derived docking pose of diacetyl within the binding groove of LITE-1 is available at https://doi.org/10.6084/m9.figshare.32745054.